---
title: "2023: Day 3 - Gear Ratios"
date: 2023-12-3
categories:
  - TAG_1
  - TAG_2
draft: true
---

## Setup

[The original challenge](https://adventofcode.com/2023/day/3)

```{r}
library(aochelpers)
library(dplyr)
# other options: aoc_input_data_frame(), aoc_input_matrix()
input <- aoc_input_matrix(3, 2023)
head(input)
```

# TLDR; Solutions

## Part 1 ⭐

::: {.callout-danger}
### ❓  W

:::


## Part 2 ⭐⭐

::: {.callout-danger}
### ❓  

:::


# Walkthrough / Explainer

I get to a gondola that'll take me to the water source. But ofc the gondola is broken. Why would I expect any different. Gotta find the missing part. 

## Part 1

:::{.callout-exa icon=true}
Example Data
```{r}
exa <- c("467..114..",
				 "...*......",
				 "..35..633.",
				 "......#...",
				 "617*......",
				 ".....+.58.",
				 "..592.....",
				 "......755.",
				 "...$.*....",
				 ".664.598..")

# convert to match aoc matrix input
(exa.in <- matrix(unlist(strsplit(exa, split = "")), 
							nrow = length(exa), byrow = TRUE))
```

dot's don't mean anything, so make them missing.
```{r}
exa.in[exa.in=="."] <- NA
exa.in
```

We only care if the number is neighbors with a symbol, so convert all symbols to something common. 
```{r}
(ex3 <- gsub("\\D", "*", exa.in))
```





## Part 2

:::{.callout-exa icon=true}
Example Data
```{r}
exa <- 0
```



##### Session info {.appendix}

<details><summary>Toggle</summary>


```{r}
#| echo: false
sessioninfo::session_info(pkgs = "attached")
```


</details>





